Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-09-23 doc/network.tex
error

\rev{
I have a problem with N(a). Does your model allow for different number of acquaintances? But then your R_0 is no longer valid. Also, you are always assuming neighbours/acquaintances are chosen at random - please stress this more.
}
Infected individuals may contact the same susceptible individual multiple times, but only the first effective contact gives rise to infection in a given individual (after this, they are no longer susceptible).
Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average.
\swp{
To explore the effect of multiple contacts on realized generation intervals, we no longer assume that the population is homogeneous (and therefore do not rely on the classical definition of the basic reproductive number).
Instead, we assume that a disease spreads on a network;
infected individuals contact their ``acquaintances'' at random, but ``acquantanceships'' are predetermined by the network structure before the beginning of an epidemic \cite{trapman2016inferring}.
}

@@

\rev{I did not follow how you got \hat g. What is a, N(a) and f(a)? Explain more.}
\swp{Reviewer 2 was not happy with this description. Come back.}
The population-level egocentric kernel is found by integrating over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
Finally, the link between the growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

\section{Inferring generation-interval distributions from contact-tracing data}
\rev{
some of the 4 effects described also affect R whereas others don't. Be more clear to the reader. For instance, the ordinare Euler-Lotka equations is not fulfilled in network situation. I think Ball, Pellis and Trapman have investigated different reproduction numbers for network epidemics and related.
}

When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect.
While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
both local and global depletion effects reduce number of infections that occur and shorten generation intervals.

Since the right-censoring effect is a sampling bias, we typically want to correct for it. 
In contrast, spatial effects have the same effect on how the epidemic spreads as they do on observed generation intervals. 
We therefore expect that starting from observed generation intervals and correcting for the right-censoring effect, will allow us to estimate an ``effective'' generation interval that accurately reflects dynamics of spread. 
When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$. 
We will call the temporally corrected effective generation-interval distribution $\tsub{g}{eff}(\tau)$.
In a large homogeneously mixing population, the effective distribution is expected to be equivalent to the intrinsic distribution (since the egocentric and local depletion are negligible in the absence of network structure, and global depletion is negligible early in the epidemic).

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-censoring severely underestimates the reproductive number.
Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially.
Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise.
Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
}
\label{fig:cmp}
\end{figure}

@@

The likelihood of the non-homogeneous poisson process can be written as:
\begin{equation}\label{eq:nonhomogeneous}
\begin{aligned}
&\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{censor})\\
&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR \, \tsub{G}{eff}(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta) \right),
\end{aligned}
\end{equation}
where $N$ is the total number of infected individuals.
Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
Note that the estimate of $\RR$ is sensitive to under-reporting;
instead, we use the estimated generation-interval distribution to infer the effective reproductive number from the estimated growth rate (using the Euler-Lotka equation).
\cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.

>>>>

\rev{
I have a problem with $N(a)$. Does your model allow for different number of acquaintances? But then your $R_0$ is no longer valid. Also, you are always assuming neighbours/acquaintances are chosen at random - please stress this more.
}
Infected individuals may contact the same susceptible individual multiple times, but only the first effective contact gives rise to infection in a given individual (after this, they are no longer susceptible).
Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average.
\swp{
To explore the effect of multiple contacts on realized generation intervals, we no longer assume that the population is homogeneous (and therefore do not rely on the classical definition of the basic reproductive number).
Instead, we assume that a disease spreads on a network;
infected individuals contact their ``acquaintances'' at random, but ``acquantanceships'' are predetermined by the network structure before the beginning of an epidemic \cite{trapman2016inferring}.
}

@@

\rev{I did not follow how you got $\hat g$. What is a, $N(a)$ and $f(a)$? Explain more.}
\swp{Reviewer 2 was not happy with this description. Come back.}
The population-level egocentric kernel is found by integrating over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
Finally, the link between the growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

\section{Inferring generation-interval distributions from contact-tracing data}
\rev{
some of the 4 effects described also affect R whereas others don't. Be more clear to the reader. For instance, the ordinare Euler-Lotka equations is not fulfilled in network situation. I think Ball, Pellis and Trapman have investigated different reproduction numbers for network epidemics and related.
}
\swp{They all affect R. I guess we weren't clear enough. Also, the point of our paper is that we can use 
Euler-Lotka equations as long as we know which distribution to use. I tried to clarify these points.}

When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect;
\swp{we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number.}
While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
both local and global depletion effects also reduce number of infections that occur and shorten generation intervals.

Since the right-censoring effect is a sampling bias, we typically want to correct for it. 
In contrast, spatial effects have the same effect on how the epidemic spreads as they do on observed generation intervals. 
We therefore expect that starting from observed generation intervals and correcting for the right-censoring effect, will allow us to estimate an ``effective'' generation interval that accurately reflects dynamics of spread. 
When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$ \swp{through the Euler-Lotka equation}. 
We will call the temporally corrected effective generation-interval distribution $\tsub{g}{eff}(\tau)$.
In a large homogeneously mixing population, the effective distribution is expected to be equivalent to the intrinsic distribution (since the egocentric and local depletion are negligible in the absence of network structure, and global depletion is negligible early in the epidemic).

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-censoring severely underestimates the reproductive number.
Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially.
Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise.
\swp{Here, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.}
Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
}
\label{fig:cmp}
\end{figure}

@@

The likelihood of the non-homogeneous poisson process can be written as:
\begin{equation}\label{eq:nonhomogeneous}
\begin{aligned}
&\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{censor})\\
&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR \, \tsub{G}{eff}(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta) \right),
\end{aligned}
\end{equation}
where $N$ is the total number of infected individuals.
Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
We expect estimates of $\RR$ to be sensitive to under-reporting;
instead, we use the estimated generation-interval distribution to infer the effective reproductive number from the estimated growth rate (using the Euler-Lotka equation).
White and Pagano \cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.
